Is Vagueness Beneficial for Hearers?\\Evidence from Experiments

### 2016-09-17 papers/2016-Special-Issue/v-special-issue-2016.tex
new structure

\documentclass[doc,apacite]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}
\usepackage{rotating}
\title{Vagueness as cost reduction}
\date{\today}
\authornote{\today ~\currenttime}
\shorttitle{Vagueness as cost reduction}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language is vague, yet standard game-theoretic models make it difficult to see how vague expressions can have benefits over crisp ones in co-operative situations. We report a series of experiments that aims to separate the utility of vagueness from the utility of other factors that come together with vagueness. We argue that the results support a view of vagueness where the benefits that vague terms exert are due to other influences that vagueness brings with it rather than to vagueness itself.}

\begin{document}
\maketitle               
\section{Introduction}
Vagueness pervades the language that we use on a daily basis. In everyday use, language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.} In most academic use though, the word `vagueness' has a more particular meaning. Keefe and Smith, for example, state ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \cite[p.\ 4]{keefe1996vagueness} (a similar definition can be found in \citeA{EgreKlinedinst}, among many others).  The most crucial of these criteria is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \cite[p.\ 1]{lipmanvague}. A typical example is the word ``tall'', as applied to people for example, because here is no precise, known height which separates those who are tall from those who are not. The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

Linguists, philosophers of language, and more recently game theorists, have asked why natural languages contain so many vague expressions, which are used so frequently \cite{Lipman:2000fk, lipmanvague}. By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \cite<>[p.~1]{lipmanvague}. Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport. Lipman argues that, in such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall''). Lipman uses this scenario to explain why standard game theory models of communication \cite<e.g.,>{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs. The relevant conditions are, broadly speaking, that both interlocutors know all the relevant facts (e.g., both know the person's height precisely) and that the setting for the communication is co-operative. These conditions exclude deliberate deception, and rhetorical situations like political debate and advertising where the intention is to persuade the interlocutor to adopt some point of view, where the persuasion might not be intended to be in the addressee's best interest.

Lipman argued that such the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.

Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, van2010vagueness}. Prominent among these answers is the idea that vague expressions are somehow easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this characterisation of the utility of vague language as the \emph{cost reduction} hypothesis. The idea of vagueness as cost reduction can take various shapes, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). This predicts that whenever a vague expression can perform the same communicative task as a precise one, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

Questions concerning optimal language use have many practical applications. Natural Language Generation\footnote{NLG systems take data or formulas as input, and transform them into natural language outputs \cite<>[]{reiter2000building}. The process parallels language production in humans.}  (henceforth NLG) systems must make decisions between different formulations of the same information. For example, if a man's height is 6 foot 2 inches, this could be expressed as ``187.96 metres'', ``6 foot 2'', or ``tall'', among other ways, and the NLG system must decide between these.  The problem is particularly relevant for NLG systems that take numbers as input, as many do. In the context of an NLG system faced with a practical decision of this kind, Lipman's question becomes ``Under what circumstances should vague terms be produced?''  Relevant applications include weather forecasting on the basis of numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurement such as oxygen saturation, heart rhythm, etc. \cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. At present, such NLG systems are often forced to make decisions concerning the level of precision in the utterances that they generate (e.g., ``the temperature will be in the high twenties tomorrow'') on the basis of little more than intuition. A better understanding of the benefits of different precision levels for readers would allow these systems to become more useful.

The cost reduction hypothesis is of direct relevance to psycholinguists interested in language comprehension, and additionally to psycholinguists interested in language production, for example in connection with the question of audience design \cite{Clark1982287}. For, to the extent that speakers and writers choose vague expressions over and above crisp ones because the former are easier to process for hearers than the latter, the cost reduction hypothesis suggests that speakers design their utterance for optimal benefit to their hearers -- out of altruism, so to speak.

The utility of vagueness is the attested aim of a small number of studies, but most of these have focussed on vagueness in a different sense, and focussing on different types of benefits for hearers. Two recent studies can illustrate both issues.

@@

In using references to quantities to test the cost reduction hypothesis we are only testing one aspect of vagueness in a particular context. This limits the applicability of our results. It also has the advantage that we can explore the costs and benefits of vagueness more thoroughly in that context. Since one prevalent view of vagueness is that a vague expression is never preferable to a crisp equivalent, a demonstration of a benefit for vagueness in any context would advance the discussion.

@@

Experiment 1 showed us that responses were faster and more accurate when the instructions were vague than when they were crisp, but the experiment could not distinguish effects of vagueness from those of number-avoidance or selection task: the vague conditions were also in verbal rather than numerical format; and mandated a comparison strategy rather than a matching strategy.  Experiment 2 showed us that number-avoidance in the verbal format instructions is an important factor driving the faster response times in the task, and that vagueness does not have any additional explanatory power in either the verbal format instructions or the numerical format instructions when we generated verbal and numerical versions of both crisp and vague instructions. However the experiment could not distinguish benefits of number-avoidance in the verbal instructions from benefits of the comparison selection task: the verbal instructions also mandated a comparison strategy rather than a matching strategy. In experiments 3 and 4 we manipulated vagueness and the selection task independently of numerical format. We found that there are effects of the selection task mandated by the instruction, with the comparison task instructions attracting faster response times than the matching instructions, and that vagueness exerts benefits when the selection task is \emph{comparison}, but not when the task is \emph{matching}.

@@

A comparison may clarify the logic of the situation. In recent years a number of studies, focussing typically on red wine, have suggested that alcohol, consumed in low doses, may have health benefits. An alternative explanation, however, asserts that it is not the alcohol in the wine that is beneficial, but the antioxidants from grapes. If this alternative explanation is correct then alcohol may not be as beneficial as some of us would have it. % KvD Good hey?

\bibliography{vagueness-dec-2015}
\clearpage
\appendix

>>>>

\documentclass[doc,apacite]{apa6}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{color}
\usepackage{booktabs}
\usepackage{layout}
\usepackage{apacite}
\usepackage{datetime}
\usepackage{rotating}
\title{Is Vagueness Beneficial for Hearers?\\Evidence from Experiments}
\date{\today}
\authornote{\today ~\currenttime}
\shorttitle{Is Vagueness Beneficial for Hearers?}
\twoauthors{Matt Green}{Kees van Deemter}
\twoaffiliations{Computing Science\\ University of Aberdeen}{Computing Science\\ University of Aberdeen}

\abstract{Much of everyday language is vague, but the causes of this phenomenon are not well understood. Consequently, it is difficult for the designers of a Natural Language Generation (NLG) system to know when and how to let the system generate vague expressions. The present paper is an attempt to find out what benefits vagueness might have for readers.% , by means of a series of carefully controlled psycholinguistic experiments that focus on the use of vagueness in referring expressions. The work is motivated by NLG but has implications for linguistics, philosophical logic and game theory. 
%
The paper reports on a series of experiments that aim to separate the utility of vagueness (as defined by the existence of borderline cases) from the utility of other factors that tend to co-occur with vagueness. We argue that the evidence so far supports a view of vagueness where the benefits that vague terms exert are due to other influences, rather than to vagueness itself. These factors include: low granularity; the use of evaluative words; the avoidance of overtly numerical words; the existence of comparison strategies; and, lastly and more tentatively, a phenomenon that we call range reduction. Although it is possible that other types of vague expressions (i.e., vague words outside referential noun phrases) behave differently, our work suggests that vagueness itself may not increase the utility of an expression. The paper concludes with a brief discussion of the implications of our work for practical NLG.}

\begin{document}
\maketitle               
\section{Introduction}\label{introduction}
Vagueness pervades the language that we use on a daily basis, and the challenge of understanding vague language has been a prominent concern in many areas of logic and linguistics, involving both theoretical and applied work, including the area known as Natural Language Generation (NLG).

NLG systems transform data and formulas into language \cite<e.g.,>[]{reiter2000building}. NLG systems routinely make decisions between different formulations of the same information. For example, if the temperature is 27.2 degrees Celsius, this could be expressed as ``27.2 degrees'', ``approximately 27 degrees', ``above 25 degrees", or ``warm'', and the system must decide between these.  The problem is especially important for NLG systems that take numbers as input, for example in the generation of textual weather reports from numerical weather data such as temperature and wind speed \cite{goldberg1994using,turner2006generating}, and medical decision support on the basis of clinical measurement such as oxygen saturation, heart rhythm, etc. \cite{Hripcsak01032009, hunter2008summarising, portet2009automatic}. Such systems are often forced to make decisions concerning the level of precision in the utterances that they generate on the basis of little more than intuition. Even when NLG systems are designed to mimic human language use (e.g., Konstas and Lapata 2013), 
% Ioannis Konstas and Mirella Lapata. 2013. A global model for concept-to-text generation. J. Artif. Intell. Res. (JAIR), 48:305Ð346. 
there is no guarantee that these decisions taken by these systems benefit readers. A better understanding of the benefits (for readers and hearers) of different precision levels would allow these systems to become more useful. The present article investigates the benefits,  or de-benefits, of vagueness.

%. In the context of an NLG system faced with a practical decision of this kind, Lipman's question becomes ``Under what circumstances should vague terms be produced?''  Relevant applications include weather forecasting

Language use may be called vague for various reasons.\footnote{See e.g. the entry ``vague'' in \cite{penguin}.} In most academic use though, the word `vagueness' has a specific meaning. Keefe and Smith, for example, state ``vague predicates have borderline cases, have fuzzy boundaries, and are susceptible to sorites paradoxes'' \cite[p.\ 4]{keefe1996vagueness}, also \citeA{EgreKlinedinst}).  The crucial criterion is the existence of borderline cases: ``a word is precise if it describes a well-defined set of objects. By contrast, a word is vague if it is not precise'' \cite[p.\ 1]{lipmanvague}. A typical example is the word ``tall'', as applied to people for example, because here is no precise, known height which separates those who are tall from those who are not. The crucial point is that ``tall'' admits borderline cases (i.e., people who may or may not count as tall), which are the hallmark of vagueness as we use the term.

Linguists, philosophers of language, and more recently game theorists, have asked why natural languages contain so many vague expressions \cite{Lipman:2000fk, lipmanvague}. By introducing borderline cases, these expressions create potential misunderstandings, thereby creating ``a worldwide several-thousand year efficiency loss'' \cite<>[p.~1]{lipmanvague}. Lipman explains the point by means of a scenario in which a speaker describes a person to a hearer, who needs to identify that person in the arrivals hall of of an airport. In such a scenario, a precise description of the person's height (e.g., ``The person's height is 187.96 cm'') would be more useful than a vague one (``The person is tall''). Lipman uses this scenario to explain why standard game theory models of communication \cite<e.g.,>{Crawford:1982lr} predict that, under certain conditions, a crisp act of communication will always have more utility than a vague act that communicates the same state of affairs. %The relevant conditions are, broadly speaking, that both interlocutors know all the relevant facts (e.g., both know the person's height precisely) and that the setting for the communication is co-operative. These conditions exclude deliberate deception, and rhetorical situations like political debate and advertising where the intention is to persuade the interlocutor to adopt some point of view, where the persuasion might not be intended to be in the addressee's best interest.

Lipman argued that the efficiency loss resulting from vague expressions would be unlikely to have arisen unless there are advantages as well as disadvantages associated with vague expressions. Lipman asked, essentially, what these advantages might be. %, and how they might find a place in a game-theoretical explanation. In this paper, we focus on the first part of Lipman's question.
Several tentative answers to Lipman's question have been offered  \cite<see>{van2009utility, van2010vagueness}. Prominent among these answers is the idea that vague expressions are easier to process, by a speaker and/or a hearer, than expressions that are not vague (i.e., crisp) \cite<e.g.,>[]{lipmanvague,De-Jaegher:2003lr,vanrooij2003lr}. For example, \citeA<>[p.\ 11]{lipmanvague} writes: ``For the listener, information which is too specific may require more effort to analyze''. We shall refer to this as the \emph{cost reduction} hypothesis. %The idea of vagueness as cost reduction can take various shapes, but the basic hypothesis is that it is easier for people to think in terms of loosely defined categories (such as ``quite a few'', or ``many'') than in terms of crisply defined ones (such as ``thirteen'', or ``237''). 
%This hypothesis predicts that when a vague expression can perform the same communicative task as a precise one, then, the things being equal, it is rational to choose the vague expression. The corollary for comprehension is that vague expressions should be understood more readily than precise expressions.

%The cost reduction hypothesis is of direct relevance to psycholinguists interested in language comprehension, and additionally to psycholinguists interested in language production, for example in connection with the question of audience design \cite{Clark1982287}. For, to the extent that speakers and writers choose vague expressions over and above crisp ones because the former are easier to process for hearers than the latter, the cost reduction hypothesis suggests that speakers design their utterance for optimal benefit to their hearers -- out of altruism, so to speak.

Charting the utility of vagueness is the attested aim of a small number of studies, but most of these have focussed on vagueness in a different sense, and focussing on different types of benefits for hearers. Two recent studies can illustrate both issues.

@@

In using references to quantities to test the cost reduction hypothesis we are only testing one aspect of vagueness in a particular context. This limits the applicability of our results. However, it has the advantage that it enables us to explore the costs and benefits of vagueness more thoroughly. Since one prevalent view of vagueness is that a vague expression is never preferable to a crisp equivalent, a demonstration of a benefit for vagueness in any context would advance the discussion.

@@

Experiment 1 showed us that responses were faster and more accurate when the instructions were vague than when they were crisp, but the experiment could not distinguish effects of vagueness from those of number-avoidance or selection task: the vague conditions were also in verbal rather than numerical format; and mandated a comparison strategy rather than a matching strategy.  Experiment 2 showed us that number avoidance in the verbal format instructions is an important factor driving the faster response times in the task, and that vagueness does not have any additional explanatory power in either the verbal format instructions or the numerical format instructions when we generated verbal and numerical versions of both crisp and vague instructions. However the experiment could not distinguish benefits of number-avoidance in the verbal instructions from benefits of the comparison selection task: the verbal instructions also mandated a comparison strategy rather than a matching strategy. In experiments 3 and 4 we manipulated vagueness and the selection task independently of numerical format. We found that there are effects of the selection task mandated by the instruction, with the comparison task instructions attracting faster response times than the matching instructions, and that vagueness exerts benefits when the selection task is \emph{comparison}, but not when the task is \emph{matching}.

@@

A comparison may clarify the logic of the situation. In recent years a number of studies, focussing typically on red wine, have suggested that alcohol, consumed in low doses, may have health benefits. An alternative explanation, however, asserts that it is not the alcohol in the wine that was beneficial, but antioxidants from grapes. If this alternative explanation is correct then alcohol may not be as beneficial as some would like to think.\\[2ex]
%
{\bf Implications for practical NLG systems.} Our findings suggest a re-think of the questions on which much research on the utility of vagueness rests. The question of how a particular piece of quantitative information is best conveyed through language is certainly an important one for practical NLG. Years of research on the logic of vagueness in natural language have primed the research community to expect that the utility of vagueness is an important part of the answer, but our findings call this into question. %It would appear that this is true for most vague expressions, and not only the ones investigated in our experiments, where a speaker is referring to a square with dots.

Consider, once again, a number of ways in which a given temperature can be reported. Given a temperature measurement (or prediction) as input, an NLG system might justifiably say that it was
%
\begin{quote}
(a) {\em 27.2 degrees}, or\\
(b) {\em approximately 27 degrees}, or\\
(c) {\em above 25 degrees}, or\\
(d) {\em warm}, 
\end{quote}
%
among other candidate expressions. If standard accounts are to be believed, then options (a) and (c) convey crisp information, whereas (b) and (d) are vague (i.e., they permit borderline cases). But, to the best of our knowledge, there is no experimental evidence that the fact in itself that an expression is vague confers a benefit (or a de-benefit) on it for hearers. Rather than asking whether a candidate expression is vague, other questions might shed more light on the NLG system's choice, similar (though not necessarily identical) to the ones identified in our empirical studies. For a start, both (c) and (d) designate an interval that is half-bounded, whereas both (a) and (b) express closed intervals, and this might affect their comprehension by a hearer (cf., our finding about comparison versus matching strategies). Furthermore, the expressions (c) and (d) -- one of which is vague while the other is crisp -- appear to convey very similar amounts of information, saying that the temperature is higher than some (fairly high) standard. Perhaps most importantly, "warm" is shorter than the other three expressions and avoids the use of numbers, and our experiments suggest that this may make ``warm" more rapidly understood than its competitors; earlier experiments point in the same direction, given the evaluative nature of the word ``warm" (see section 1 and \citeA{peters2009bringing}).

One way to see why vagueness may not matter as much for NLG -- and for human communication more generally -- as is often thought is the following thought experiment. Many speakers understand the word ``warm" as vague. Now suppose a group of users agreed to give it a precise definition; according to this precisification, ``warm" means $>25$ degrees (as in (c)). It seems unlikely that this change, from a vague meaning to a crisp one, would change the utility of the word.

\bibliography{v-special-issue-2016}
\clearpage
\appendix
